---
title: Preferred medications for opioid agonist therapy and associated factors among people who regularly use opioids in Australia
author:
  - name: MJ Stowe
    orcid: 0000-0002-0760-5497
    corresponding: true
    email: mjstowe@kirby.unsw.edu.au
    roles:
      - Investigation
      - Project administration
      - Software
      - Visualization
    affiliations:
      - The Kirby Institute, UNSW Syndey, Sydney Australia
  - name: Anna Conway
    orcid: 0000-0002-7859-8394
    corresponding: false
    roles: []
    affiliations:
      - The Kirby Institute, UNSW Syndey, Sydney Australia
  - name: Frederick Altice
    orcid: 0000-0002-7859-8394
    corresponding: false
    roles: []
    affiliations:
      - Yale School of Medicine, Yale University, New Haven, Connecticut, United States
  - name: Tom Hassett
    orcid: 0000-0002-7859-8394
    corresponding: false
    roles: []
    affiliations:
      - Yale School of Medicine, Yale University, New Haven, Connecticut, United States
  - name: Marianne Byrne
    orcid: 0000-0002-7859-8394
    corresponding: false
    roles: []
    affiliations:
      - The National Drug and Alcohol Research Centre, UNSW Syndey, Sydney, Australia
      - The Kirby Institute, UNSW Syndey, Sydney Australia
  - name: Samantha Colledge
    orcid: 0000-0002-7859-8394
    corresponding: false
    roles: []
    affiliations:
      - The National Drug and Alcohol Research Centre, UNSW Syndey, Sydney, Australia
  - name: Phillip Read
    orcid: 0000-0002-7859-8394
    corresponding: false
    roles: []
    affiliations:
      - Kirketon Road Centre, Sydney, New South Wales, Australia 
      - Drug and Alcohol Services, Northern Sydney Local Health District, Sydney, New South Wales, Australia
  - name: Mark Montebello
    orcid: 0000-0002-7859-8394
    corresponding: false
    roles: []
    affiliations:
      - The National Drug and Alcohol Research Centre, UNSW Sydney, UNSW, Australia
      - Drug and Alcohol Services, Northern Sydney Local Health District, Sydney, New South Wales, Australia
      - University of Sydney, Sydney, New South Wales, Australia
  - name: Jeremy Hayllar
    orcid: 0000-0002-7859-8394
    corresponding: false
    roles: []
    affiliations:
      - Drug Health Service, South West Sydney Local Health District, New South Wales, Australia
  - name: Sione Crawford
    orcid: 0000-0002-7859-8394
    corresponding: false
    roles: []
    affiliations:
      - Harm Reduction Victoria, Victoria, Australia
  - name: Charles Henderson
    orcid: 0000-0002-7859-8394
    corresponding: false
    roles: []
    affiliations:
      - New South Wales Users and AIDS Association (NUAA), Sydney, Australia
  - name: Michael Farrell
    orcid: 0000-0002-7859-8394
    corresponding: false
    roles: []
    affiliations:
      - The National Drug and Alcohol Research Centre, UNSW Sydney, Sydney, Australia
  - name: Chris Gough
    orcid: 0000-0002-7859-8394
    corresponding: false
    roles: []
    affiliations:
      - CAHMA
  - name: Louisa Degenhardt
    orcid: 0000-0002-7859-8394
    corresponding: false
    roles: []
    affiliations:
      - The National Drug and Alcohol Research Centre, UNSW Syndey, Sydney, Australia
  - name: Amy Peacock
    orcid: 0000-0002-7859-8394
    corresponding: false
    roles: []
    affiliations:
      - The National Drug and Alcohol Research Centre, UNSW Syndey, Sydney, Australia
  - name: Gregory J. Dore
    orcid: 0000-0002-7859-8394
    corresponding: false
    roles: []
    affiliations:
      - The Kirby Institute, UNSW Syndey, Sydney Australia
  - name: Jason Grebely
    orcid: 0000-0002-7859-8394
    corresponding: false
    roles: []
    affiliations:
      - The Kirby Institute, UNSW Sydney, Australia
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  html: 
    toc: true
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
always_allow_html: true
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
# bibliography: references.bib
# csl: "../templates/international-journal-of-drug-policy.csl" # Insert path for the bib-style
abstract: |
    **Background:** People who use opioids from an unregulated drug supply are increasingly exposed to related adverse social and health outcomes. Opioid agonist therapy (OAT) is effective in mitigating adverse outcomes, with individual preference identified as key to enhancing treatment. We examined OAT preferences and associated factors in a national sample of people who regularly use opioids. **Methods:** In this cross-sectional study, 400 people were enrolled between October 2020-April 2021. Participants completed an interviewer-administered questionnaire with questions on sociodemographic, behavioural, drug use- and treatment-related characteristics. Multivariable logistic regression was used to evaluate factors associated with treatment preferences. **Results:** Among all participants (mean age 45, 42% female), X%  preferred OAT, with X% and X% reporting a preference for methadone and buprenorphine respectively. Among those who reported a preference for OAT, X% (n=X) preferred metahdone, while X% (n=X) preferred buprenorphine. participants who preferred methadone were more likely to have ever received OAT with methadone, and had used heroin (aOR X.XX, 95%CI:X.XX, X.XX) and other pharmaceutical opioids in the past month(aOR X.XX, 95%CI:X.XX, X.XX). Participants who preferred buprenorphine were more likely to have ever received OAT with burpenorphine (aOR X.XX, 95%CI:X.XX, X.XX). Among participants currently on OAT (n=317), X% in receipt of methadone and X% of buprenorphine were prescribed their preferred medication. Participants who preferred methadone were more likely to be receiving OAT with the same medication (aOR X.XX, 95%CI:X.XX, X.XX), had used and heroin (aOR X.XX, 95%CI:X.XX, X.XX) and other pharmaceutical opioids in the prior month (aOR X.XX, 95%CI:X.XX, X.XX). Participants who preferred buprenorphine were more likely to be currently receiving burpenorphine (aOR X.XX, 95%CI:X.XX, X.XX). **Conclusions:** Among people who regularly used opioids, most preferred to receive OAT with methadone and those currently receiving OAT were prescribed their preferred medication. Providers that do not offer a range of OAT and restrict choice of preferred medication risk negatively impacting treatment outcomes.
keywords: |
  - Opioid agonist therapy
  - Medications for opioid use disorder
  - Methadone
  - Burpenorphine
key-points: |
  - From our sample of people who regularly use opioids from around Australia. most preferred to receive opioid agonist therapy with methadone
  - Current OAT participants (n=317) comprised 78% (n=x) on methadone and 91% (n=x) on buprenorphine, with 78% (n=x) and 91% (n=x) receiving their preferred medication reservoirs 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The actual document text starts here: -->

# Introduction

People who use opioids from the unregulated market are increasingly
exposed to related social and health harms [@milaney2022;
@cheetham2022], including fatal and non-fatal opioid-involved overdose
[@santo2021; @åstrøm2023], incarceration [@gisev2019; @fazel2017],
stigma [@treloar2022] and structural violence [@lancaster2023]. Use via
injection is associated is several health risks [@brener2022;
@colledge2019; @degenhardt2019] such as increased risk of HIV and HCV
infection [@Grebely2022; @Martinello2017a; @hajarizadeh], skin and soft
tissue infection [@wheeler2022] and endocarditis [@wurcel2016;
@see2020]. Opioid agonist therapy (OAT) with methadone and buprenorphine
is effective in reducing associated harms [@degenhardt2023;
@Nielsen2016; @Nielsen2022; @Jones2022], with retention associated with
improved social and health outcomes , and reduced quality-adjusted
life-years lost compared to no treatment or non-pharmacological
interventions [@martin2022]. Despite the efficacy of these medications
[@degenhardt2023], treatment acceptability, satisfaction, and retention
is sub-optimal, globally. Individual preferences for these medications
play a pivotal role in determining positive treatment outcomes
[@joosten2008]. With similar outcomes for a range of measures among
particular subgroups between medications [@degenhardt2023], the choice
between methadone and buprenorphine should be informed by consultation
with each person after consideration of preferences and the relative
risks and benefits of each medication to the indivudal. While a number
of studies explore the role of treatment preferences generally
[@uebelacker2016; @kenney2018], relatively fewer number of studies focus
specifically on factors associated with preferences for medication types
and formulations.

Preferences around OAT and individual choice are identified as key to
improving treatment outcomes and reducing potential harms associated
with use of unregulated opioids [@muthulingam2023]. Preferences are
comprised of attitudes, beliefs, expectations, values, and processes
used to evaluate the costs and benefits of treatment options
[@montori2013], with the decision for one formulation over another being
largely preference sensitive [@keirns2009]. Understanding treatment
preference could inform shared decision-making and help to overcome
barriers to effective, person-centered healthcare and support
[@maddensatisfaction2008]. However, about how opioid-dependent patients'
preferences and previous experiences influence treatment decisions.
Preferences, perceptions and acceptability of medication formulations
vary widely based on each participant's individual experiences, goals,
and values [@neale2018; @nealequalitative2023; @nealedepot2019].
Previous studies exploring preferences for methadone and buprenorphine
have beenStudies have compared patients' views of methadone with
buprenorphine [@hillcomparison2015]. However, findings have been
inconsistent and variable [@whitepatients2007]. Previous studies into
treatment preferences carried out in the United Kingdom [@neale2018;
@tompkinsopioid2019], France [@rolland2021], North America
[@saunders2020; @muthulingam2023; @kaplowitztreatment2022], Iran
[@amini-rarani2023] and Australia [@larance2020] suggest that
preferences for OAT is influenced by a range of physical, psychological
and social factors. Additional studies have explored preferences
generally, however, understanding of factors that influence and
contribute to stated preference are limited. Limitations of previous
studies include limited geographic scope with samples from single
clinics (e.g.[@bailey2013]) or cities (e.g.[@luty2004]), and relatively
small samples and lack of power to investigate factors associated with
OAT preferences (e.g. [@ridge2009]).

In Australia, there are several formulations of buprenorphine currently
available, including a monobuprenorphine formulation (Subutex) and
buprenorphine-- naloxone formulations in a tablet or film and
administered orally or sublingually (Suboxone), and, more recently,
long-acting injectable formulations that are administered once-weekly or
once-monthly, depending on the product (Buivdal and Sublocade). Most
people received methadone (58%), the median age was 44 years. ODT
medicines were mostly prescribed by a private health practitioner,
dosing was mainly dispensed in pharmacies, and the primary drug of
concern for the majority was heroin. In 2021, approximately 47,000
people were receiving OAT (median age 44; over two-thirds male), with
most (58%) receiving methadone. The median age was 44 years and
two-thirds were male. There are variations in OAT provision between
jurisdictions due to decentralized funding of health services and the
varied historical contexts across the country [@hall2023]. In some
jurisdictions, OAT is dispensed exclusively at community pharmacies,
while others have a mix of community pharmacy and public clinics.
Community pharmacies offer longer opening hours and more accessible
locations than public clinics, yet the out-of-pocket dispensing fees at
pharmacy can make OAT prohibitively expensive [@tran2022; @zahra2022].
Historically, provision of OAT has been considered complex given its
strict regulatory oversight and varied interpretation of guidelines
among OAT prescribers. Despite providers differing in their
interpretation of guidelines, OAT provision in Australia has been
criticised for its rigidity in not adapting to the needs of people
engaged in treatment [@crawford2013]. Many aspects of OAT provision also
lack flexibility, and are often not person-centered, with limited access
to unsupervised dosing and restrictive dosing times, which cumulatively
decrease treatment adhearance [@hall2023]. It is important to understand
preferences for the available medications to inform person-centered OAT
provision and enhance treatment outcomes.

Accessibility and diversity of OAT need to be enhanced to ensure that
all people can receive their preferred OAT. Improved understanding of
preference and factors associated with a preference for OAT preference
is needed to inform policy and service provision. information on
preference for different OAT options has shown to enhance patient
acceptability, leading to increased OAT treatment uptake, adherence, and
retention to improve clinical utility and facilitate planning,
prioritization and investment in national strategies and guidelines.
Extensive literature on patient participation in medication
decision-making emphasizes that incorporating patient treatment
preference leads to higher rates of retention and the desired outcomes
[@friedrichs2016; @joosten2008]. Providers incorporating client
preferences into treatment is linked to increased satisfaction, with
individual choice, with alignment with preferred therapy leading to
improved outcomes [@fallah2015; @joosten2008]. To better understand
treatment preferences and develop an evidence-base to inform policy and
treatment service design and delivery, we examined OAT preferences and
associated factors among a national sample of people who regularly use
opioids in Australia.

# Data & Methods {#sec-data-methods}

## Study design and participants {#sec-study-design}

In this cross-sectional study, participants were enrolled from 59 sites
including drug and alcohol treatment clinics (n=28), needle and syringe
programmes (n=13), pharmacies and organisations advocating for (n=6) and
providing support and services to people who use drugs (n=7), There was
national representation of sites across all states and territories in
Australia (except Tasmania), including Australian Capital Territory
(n=3), New South Wales (n=37), the Northern Territory (n=1), Queensland
(n=9), South Australia (n=6), Victoria (n=2), and Western Australia
(n=1). Study enrolment took place between October 2020 and April 2021,
and continued throughout periods of COVID-19-related restrictions. To be
eligible for this study, participants had to be aged 18 years or older,
provide voluntary and informed consent, use opioids regularly. In this
study, regular opioid use was defined as use of any opioid (including
heroin, methadone, burpenorphine use and other opioids or the
extramedical use of pharmaceutical opioids) on at least 21 of the past
28 days. The content within this article is presented following the
'Strengthening the reporting of observational studies in epidemiology'
(STROBE) guidelines [@von2007; @Vandenbroucke2007]. A completed
checklist of the STROBE requirements is available in the
@sec-supp-material.

## Procedures {#sec-procedures}

Participants were recruited from a range of settings (including drug and
alcohol treatment services, needle and syringe programmes, organisations
providing support and services for people who use drugs, community-based
and -led advocacy organisations, and pharmacies), via snowballing (where
eligible participants promote the study via their personal networks),
and word-of-mouth. The study was advertised at services using posters
and fliers, but service staff were not directly involved in recruitment.
Participants who were interested in the study contacted the study team
directly and were screened for eligibility over the telephone.

Interviewers contacted the person over the phone or video conference
(Microsoft Teams or Zoom) to obtain consent and conduct the interview.
Interviews were semi-structured were conducted by trained interviewers.
Participants completed an interviewer-administered questionnaire focused
on patient preferences for OAT that included information on participant
demographics, drug use characteristics, and drug treatment. The
interviews took approximately 30-45 minutes to complete, and
participants were reimbursed AUD\$40 for their time and out-of-pocket
expenses.

### Outcomes {#sec-outcomes}

The primary outcome was a preference for opioid agonist therapy with
methadone, which was assessed by asking the question "*Of all the
following types of medications used for opioid agonist therapy, if you
could choose today, which one would you prefer?*" Options included
methadone, monobuprenorphine (Subutex), buprenorphine-naloxone
(Suboxone) taken orally as a film or tablet, long acting injectable
buprenorphine (Buvidal or Sublocade), any treatment (i.e. no preference
for any medication), and no treatment.

Demographic, behavioural, and clinically significant factors
hypothesised to be associated with preference to receive OAT with
methadone were determined a priori, comprising the following: 1) age at
survey (18 - 35; 36 -- 45; \> 45 years old) ), 2) gender (female; male;
transgender), 3) education (\< year 10; \> year 10), 4) employment
(paid; other), 5) homelessness (no; yes), 6) chronic pain (no; yes), 7)
incarceration history (never, \>6 months ago, \\6 months ago), 8) Drugs
used and injected in the last 28 days (heroin, non-prescribed methadone,
non-prescribed buprenorphine, other pharmaceutical opioids,
methamphetamine, cocaine and benzodiazepines), 9) Frequency of injection
drug use (\>1 year ago, within 1--12 months ago, within the last month
less than daily, and daily or more), 10) OAT (history and current), with
methadone (history and current) and buprenorphine (history and current).
Among the sub-group of participants currently receiving OAT, factors
hypothesized to be associated with a preference for OAT with methadone
included 11) current OAT medication (methadone; buprenorphine;
buprenorphine-naloxone; long-acting injectable buprenorphine; none), 12)
Drugs used while on OAT (no, heroin, non-prescribed methadone,
non-prescribed buprenorphine, other pharmaceutical opioids,
methamphetamine, cocaine and benzodiazepines); 13) Pay out of pocket for
oat treatment (no; yes); 14) Amount paid out of pocket for OAT (xx; xx),
15) Site last prescribed OAT medication was collected (pharmacy; public
clinic; private clinic; at home; other), 16) Frequency of OAT dose
collection (daily or several times per week; weekly or less frequently),
17) Distance traveled to OAT collection site, 18) Time traveled to OAT
collection site (less than 30 minutes; more than 30 minutes).

## Statistical analysis {#sec-analysis}

The proportion of people who reported preference for methadone and
factors associated with this preference was assessed. Quantitative
parameters are presented as the median and interquartile range \[IQR\].
Categorical parameters are presented as the number and percentage (n;
%). The association between the response to each variable and the
binarized category of preference for methadone was explored using
logistic regression modeling through unadjusted bivariate analysis,
providing an odds ratio and the 95 % confidence interval (OR \[95%
CI\]), and comparisons adjusted for age category and gender, providing
an adjusted OR (aOR) and the 95 % CI. A sample of those who would choose
treatment and who had a preference for methadone or any buprenorphine
forumation was selected for analysis (n=352) . Multivariable logistic
regression was used assess factors associated with preferred medication
for OAT. Unadjusted and adjusted odds ratios were derived using logistic
regression and 95% confidence intervals calculated for all variables.
For the final model, all the explanatory variables significantly
associated with the outcome in the bivariate models were included in the
multivariable model, whilst also adjusting for age and gender. The final
model was assessed through examining which contributed significantly to
model fit through backwards selection using decreasing values of Akaike
Information Criterion. The variance inflation factor was used to assess
collinearity in the final model [@obrien2007]. Individuals with missing
values were not integrated in the models. A subset of the sample
comprising of participants currently receiving OAT with methadone and
buprenorphine (including sublingual-monobuprenorphine,
buprenorphine-naloxone and long-acting injectable buprenorphine) was
selected for analysis. There was no formal sample size calculation,
however, based of similar previous surveys (e.g. [@larance2020;
@rolland2021]) we aimed to recruit 300 - 400 people.

The analytical approach received input from and was reviewed by people
who use opioids with lived-experience of injecting drug use and opioid
agonist therapy with various formulations of methadone and
buprenorphine. Advice and perspectives were provided on the inclusion of
explanatory variables and the clinical and practical significance of
identified factors.

The analysis was undertaken in using `R` version
`r paste0(version$major, "." ,version$minor)` using the following
packages: gtsummary v. 1.7.2 [@gtsummary], finalfit v. 1.0.6
[@finalfit], quarto v. 1.3 [@quarto] and rmarkdown v. 2.25 [@rmarkdown].
The code for this analysis is available on request from the
corresponding author. The data supporting the findings of this study are
available within the article and its Supporting information. The
analysis was not pre-registered and the results should be considered
exploratory.

## Study oversight {#sec-oversight}

All participants provided written informed consent before study
procedures. The study protocol were approved by the Human Research
Ethics Committees at St Vincent's Hospital, Sydney (HREC Ref:
HREC/17/SVH/113) and the Aboriginal Health and Medical Research Council
(HREC Ref: 1279/17). This study was conducted according to the
Declaration of Helsinki and International Conference on Harmonization
Good Clinical Practice (ICH/GCP) guidelines.

## Role of the Funding source {#sec-funding}

The study was funded by a research grant from X. The funders had no role
in the study design, data collection, analysis, interpretation of the
results, writing or the decision to submit the study for publication.
JG, AC, LD, and MJS had access to the raw data. The National Drug and
Alcohol Research Centre and the Kirby Institute, UNSW Sydney
collaborated to design the study, monitor study conduct, and perform the
statistical analysis. X, X and X were responsible for the decision to
submit the study for publication.

## Results {#sec-results}

# Methods

# Results

<!-- Here's some example analysis code: -->

```{r}
#| label: get-data
#| eval: false
# Note the path that we need to use to access our data files when rendering this document
my_data <- read.csv(here::here('analysis/data/raw_data/my_csv_file.csv'))
```

```{r}
#| label: fig-demo-plot
#| fig-cap: "A plot of random numbers"
plot(rnorm(10))
```

@fig-demo-plot shows how we can have a caption and cross-reference for a plot. Note that figure label and cross-references must both be prefixed with `fig-`

```{r}
#| label: demo-inline-code
#| echo: false
x <- round(pi, 2)
```

Here is an example of inline code `r x` in the middle of a sentence.

# Discussion

# Conclusion

# Acknowledgements

<!-- The following line inserts a page break  -->

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r}
#| label: colophon
#| cache: false

# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```
